---
title: Building BIDS Apps with niflowr and bidsappr
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Building BIDS Apps with niflowr and bidsappr} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The [bidsappr](https://github.com/bbuchsbaum/bidsappr) package lets you build containerized BIDS Apps in R with zero boilerplate. niflowr provides the neuroimaging operations through wrappers for FSL, ANTs, FreeSurfer, and AFNI. Together, you can build a standards-compliant BIDS App that wraps any combination of neuroimaging tools.

A BIDS App is a containerized neuroimaging pipeline that follows the [BIDS Apps specification](https://bids-apps.neuroimaging.io/). It accepts a BIDS dataset as input, runs participant-level and/or group-level analysis, and writes outputs in a structured format.

## Quick Start with ni_bids_app()

Scaffold a new BIDS App with a single function call:

```{r}
library(niflowr)

ni_bids_app("~/my_brain_app", tools = c("fsl.bet", "fsl.flirt"))
```

This creates:
- `app.R` with template participant/group functions
- `Dockerfile` with neuroimaging tools pre-installed
- `DESCRIPTION` and dependency metadata
- Example configuration

## Understanding the app.R Template

The generated `app.R` defines two functions: `participant()` for subject-level processing and `group()` for group-level analysis.

```{r}
participant <- function(ctx, frac = 0.5) {
  # ctx provides:
  # - ctx$subject: current subject ID
  # - ctx$files: list of input files for this subject
  # - ctx$out_file(): function to create output paths
  # - ctx$write_tsv(): function to write QC metrics

  for (subj in ctx$subjects) {
    t1w <- ctx$files[[subj]]$anat$T1w[[1]]

    # Run niflowr operation
    result <- ni_fsl_bet(
      in_file = t1w,
      out_file = ctx$out_file(subj, "brain.nii.gz"),
      frac = frac,
      .engine = "native"
    )

    # Log QC metrics
    ctx$write_tsv(
      subject = subj,
      out_size = file.size(result$out_file),
      exit_code = result$exit_code
    )
  }
}

group <- function(ctx) {
  # Access participant-level outputs
  # Run group statistics, generate reports, etc.
}
```

The `ctx` object is provided by bidsappr and handles all BIDS input/output conventions.

## Example: Brain Extraction App

A complete working example that runs BET on all T1w images:

```{r}
library(niflowr)
library(bidsappr)

participant <- function(ctx, frac = 0.5, robust = FALSE) {
  qc_metrics <- list()

  for (subj in ctx$subjects) {
    # Find T1w image
    t1w_files <- ctx$files[[subj]]$anat$T1w
    if (length(t1w_files) == 0) {
      warning("No T1w found for ", subj)
      next
    }

    t1w <- t1w_files[[1]]
    out_brain <- ctx$out_file(subj, "anat", "brain.nii.gz")
    out_mask <- ctx$out_file(subj, "anat", "brain_mask.nii.gz")

    # Run brain extraction
    result <- ni_fsl_bet(
      in_file = t1w,
      out_file = out_brain,
      frac = frac,
      robust = robust,
      mask = TRUE,
      .engine = "native"
    )

    # Collect QC metrics
    qc_metrics[[subj]] <- data.frame(
      subject_id = subj,
      input_size_mb = file.size(t1w) / 1e6,
      output_size_mb = file.size(out_brain) / 1e6,
      frac_param = frac,
      exit_status = result$exit_code,
      timestamp = Sys.time()
    )
  }

  # Write QC report
  qc_df <- do.call(rbind, qc_metrics)
  ctx$write_tsv(qc_df, "bet_qc.tsv")
}

# Run the app
if (!interactive()) {
  bidsappr::main(app_dir = "my_brain_app")
}
```

Run from the command line:
```bash
Rscript app.R /data/bids /data/output participant --frac 0.3 --robust
```

## Example: Multi-Step Preprocessing App

A more complex app that chains multiple operations:

```{r}
participant <- function(ctx, template = "MNI152_T1_2mm.nii.gz") {
  for (subj in ctx$subjects) {
    t1w <- ctx$files[[subj]]$anat$T1w[[1]]

    # Step 1: Skull strip
    brain <- ctx$out_file(subj, "anat", "brain.nii.gz")
    ni_fsl_bet(
      in_file = t1w,
      out_file = brain,
      frac = 0.5,
      .engine = "native"
    )

    # Step 2: Register to template
    mat_file <- ctx$out_file(subj, "anat", "brain_to_mni.mat")
    reg_result <- ni_fsl_flirt(
      in_file = brain,
      ref = template,
      out = ctx$out_file(subj, "anat", "brain_mni.nii.gz"),
      omat = mat_file,
      .engine = "native"
    )

    # Step 3: Apply transformation to original T1w
    ni_fsl_flirt(
      in_file = t1w,
      ref = template,
      init = mat_file,
      applyxfm = TRUE,
      out = ctx$out_file(subj, "anat", "T1w_mni.nii.gz"),
      .engine = "native"
    )

    # Log pipeline completion
    ctx$write_tsv(
      subject = subj,
      pipeline_complete = TRUE,
      registration_exit = reg_result$exit_code
    )
  }
}
```

## Container Deployment

bidsappr scaffolds a `Dockerfile` that:
- Installs R and required packages
- Pre-installs neuroimaging tools (FSL, ANTs, etc.)
- Configures the BIDS App entrypoint

```dockerfile
FROM rocker/r-ver:4.3.0

# Install neuroimaging tools
RUN apt-get update && apt-get install -y \
    fsl-core \
    ants

# Install R packages
RUN R -e "install.packages(c('niflowr', 'bidsappr'))"

# Copy app
COPY app.R /app/
WORKDIR /app

ENTRYPOINT ["Rscript", "app.R"]
```

Build and run:
```bash
docker build -t myapp .
docker run myapp /bids /output participant --frac 0.3
```

Inside the container, use `.engine = "native"` since tools are installed in the container image.

## Custom CLI Parameters

bidsappr automatically generates CLI flags from your function signatures. Function parameters become command-line options:

```{r}
participant <- function(ctx, frac = 0.5, robust = FALSE, threshold = 0.1) {
  # frac, robust, threshold are now CLI parameters
}
```

Users can then run:
```bash
Rscript app.R /bids /output participant \
  --frac 0.3 \
  --robust \
  --threshold 0.05
```

This makes niflowr pipelines instantly accessible as command-line tools without manual argument parsing.

## Best Practices

1. **Always use `.engine = "native"`** inside BIDS Apps — tools are containerized
2. **Write QC metrics** with `ctx$write_tsv()` for reproducibility
3. **Check file existence** before processing (not all subjects have all modalities)
4. **Use ctx$out_file()** to ensure outputs follow BIDS derivatives structure
5. **Keep participant() pure** — no global state, all inputs via ctx or parameters

## Next Steps

- Explore the [bidsappr documentation](https://github.com/bbuchsbaum/bidsappr) for advanced features
- Add group-level analysis in the `group()` function
- Integrate quality control dashboards
- Publish your app on [BIDS Apps](https://bids-apps.neuroimaging.io/)
