---
title: 'Extending niflowr: wrap a new tool'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Extending niflowr: wrap a new tool} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(niflowr)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Eventually you will hit a tool niflowr does not ship a wrapper for —
yesterday's preprint, your lab's in-house segmenter, a flag the upstream
spec never bothered to expose. Adding it is small work: write a JSON spec,
call a generator, ship a wrapper. This vignette walks through the loop
end-to-end using FreeSurfer's SynthStrip as the worked example.

## The mental model

Every wrapper in niflowr is built from three artefacts:

| Artefact | Where it lives | Who writes it |
|:---------|:---------------|:--------------|
| **Spec** — a JSON description of inputs, outputs, and CLI rendering | `inst/specs/<id>.json` | You |
| **Wrapper** — an `ni_*()` R function | `R/wrappers_auto.R` | `tools/gen_wrappers.R` |
| **Profile** — how the binary is run (native / docker / apptainer) | `niflowr.yml` | The user, at project level |

The spec is the source of truth. Wrappers and golden command-line
fixtures are regenerated from it. Profiles are kept separate so that the
same spec can run natively in CI, in a Docker container on a laptop, and
under Apptainer on a cluster — without changes to the spec itself.

## First win: a 12-line spec

Let's wrap `mri_synthstrip` with the bare minimum: the input volume and
the brain-mask output. We'll write the spec to a tempfile so the vignette
can run end-to-end without touching the package source.

```{r write-min-spec}
spec_path <- tempfile(fileext = ".json")
writeLines(c(
  '{',
  '  "spec_version": "0.1.0",',
  '  "id": "demo.synthstrip",',
  '  "title": "SynthStrip (demo)",',
  '  "command": "mri_synthstrip",',
  '  "inputs": {',
  '    "in_file":   {"type": "file", "required": true,',
  '                  "cli": {"argstr": "-i %s"}},',
  '    "mask_file": {"type": "file",',
  '                  "cli": {"argstr": "-m %s"}}',
  '  },',
  '  "outputs": {',
  '    "mask_file": {"type": "file",',
  '                  "path": {"from_input": "mask_file"},',
  '                  "must_exist": false}',
  '  },',
  '  "runtime": {"profile": "freesurfer"}',
  '}'
), spec_path)
```

Now load it and build a call:

```{r min-call}
spec <- ni_spec_read(spec_path)

call <- ni_call(spec,
  in_file   = system.file("DESCRIPTION", package = "niflowr"),
  mask_file = "/tmp/T1_mask.nii.gz",
  .engine   = "native"
)

ni_cmd(call)
```

That's it. The spec file plus `ni_spec_read()` plus `ni_call()` is enough
to render a safe, validated command-line vector. No shell strings, no
`paste0` — niflowr keeps the command and its arguments separate so they
can never collide with shell metacharacters.

A `dry_run` confirms the wiring without invoking the binary:

```{r min-dry-run}
ni_run(call, dry_run = TRUE, echo = FALSE)
```

## Anatomy of a spec

The minimal spec hides most of the moving parts. Here is what an input
definition can carry:

```json
"frac": {
  "type": "double",
  "desc": "Fractional intensity threshold",
  "default": 0.5,
  "validate": {"min": 0, "max": 1},
  "cli": {"argstr": "-f %.2f"},
  "constraints": {"xor": ["robust"]}
}
```

The four blocks that matter:

- **`type`** — one of `file`, `dir`, `string`, `int`, `double`, `bool`,
  `flag`, `enum`, `list`. Drives R coercion and the wrapper's `@param`
  documentation.
- **`validate`** — fail fast at `ni_call()`-time. `exists` (file/dir
  must be present), `min`/`max` (numeric range), `regex` (string
  pattern). Errors here surface as clean `cli::cli_abort()` messages
  instead of mysterious crashes inside the binary.
- **`cli.argstr`** — a `printf`-style format string. `%s` for files /
  strings, `%d` for ints, `%.2f` etc. for doubles. Flags use a literal
  switch like `-g` (no `%`).
- **`constraints`** — `xor` (mutually exclusive) and `requires`
  (co-required). Validated at call time.

### Positional vs. switched arguments

Most CLIs mix the two. niflowr handles both with the same shape:

```json
"in_file":  {"type": "file", "required": true,
             "cli": {"argstr": "%s", "position": 0}},
"out_file": {"type": "file",
             "cli": {"argstr": "%s", "position": 1}},
"frac":     {"type": "double",
             "cli": {"argstr": "-f %.2f"}}
```

`position` orders positional args (negative values count from the end —
useful for trailing arguments like `mri_watershed in.mgz out.mgz`).
Switched arguments render in alphabetical order before positionals, so
the shape is deterministic and golden-testable.

### Output paths come from inputs

Outputs declare *where* the artefact lands so downstream code (targets,
BIDS-aware tools, `ni_result$outputs`) can find it:

| Pattern | Use when |
|:--------|:---------|
| `{"from_input": "out_file"}` | the user passes the path explicitly |
| `{"template": "{sd}/{sid}/mri/aseg.mgz"}` | the path is computed from other inputs (glue interpolation) |
| `{"static": "results.txt"}` | the tool always writes to a fixed name in cwd |

Set `must_exist: false` when the output is optional (e.g. SynthStrip's
`-m` and `-d` only produce files when the user requested them).

### Runtime profile

```json
"runtime": {"profile": "freesurfer"}
```

The profile name is a *contract*, not a binding. Users wire profile names
to docker images / apptainer SIFs / native paths in their own
`niflowr.yml`. Pick a profile name that matches the binary's natural
home: `freesurfer`, `fsl`, `ants`, `afni`, or a new one. If you invent a
new profile, document the docker image users will likely point it at.

## Object zoo

Three classes carry your call through the system:

| Class | Constructor | What it holds |
|:------|:------------|:--------------|
| `ni_spec` | `ni_spec_read()` | Parsed spec — inputs, outputs, command, runtime |
| `ni_call` | `ni_call()` | Spec + bound input values, validated, ready to render |
| `ni_result` | `ni_run()` | Exit status, captured stdout/stderr, resolved output paths, provenance |

You'll write specs and read results. The `ni_call` in the middle is
what `dry_run`, `ni_cmd()`, and the golden-fixture harness all
operate on.

## Promoting a spec from local to bundled

Working from a tempfile is fine for one-off experiments. To ship the
wrapper as part of a package (your fork of niflowr, or a downstream
package that depends on it), promote the spec through the package
development loop:

**1. Drop the spec in `inst/specs/`.** Naming convention is
`<profile>.<tool>.json`, with the spec `id` matching the filename:

```
inst/specs/freesurfer.synthstrip.json
```

**2. Lint it.** This catches shell metacharacters, positional
collisions, malformed flags, and constraint references that point at
inputs that don't exist:

```{r lint, eval = FALSE}
ni_lint_specs(strict = TRUE)
```

In `fix = TRUE` mode the linter rewrites the safe cases in place
(`> %s` → `stdout_to`, dropped duplicate positions, normalised
constraint arrays).

**3. Regenerate the wrapper.** This produces `R/wrappers_auto.R` from
*all* specs in `inst/specs/` — your new one and the existing 300+:

```sh
Rscript tools/gen_wrappers.R
```

A wrapper named `ni_freesurfer_synthstrip()` now exists, with one
`@param` per input and `dry_run` / `echo` / `.cwd` / `.engine` /
`.profile` controls.

**4. Update NAMESPACE and man pages.**

```sh
Rscript -e 'roxygen2::roxygenise(".")'
```

**5. Refresh the golden command-line fixture.** This is the regression
test that protects against accidental CLI changes — every spec gets a
deterministic input/output snapshot:

```sh
Rscript tools/gen_golden_cmdline.R
```

**6. Write a test.** Mirror the pattern from
`tests/testthat/test-fastsurfer.R` or
`tests/testthat/test-freesurfer-synthstrip.R`: load the spec, build a
call with synthetic temp files, assert the rendered `cmd$args`, and
finish with a `dry_run` smoke test.

```{r test-pattern, eval = FALSE}
test_that("freesurfer.synthstrip renders core CLI args", {
  in_file <- withr::local_tempfile(fileext = ".nii.gz")
  file.create(in_file)
  call <- ni_call("freesurfer.synthstrip",
    in_file = in_file,
    mask_file = "/tmp/mask.nii.gz",
    gpu = TRUE
  )
  cmd <- ni_cmd(call)
  expect_equal(cmd$command, "mri_synthstrip")
  expect_true(all(c("-i", in_file) %in% cmd$args))
  expect_true(all(c("-m", "/tmp/mask.nii.gz") %in% cmd$args))
  expect_true("-g" %in% cmd$args)
})
```

That's the entire loop. The bundled
`inst/specs/freesurfer.synthstrip.json` is the full version of the
example we built above — about 70 lines, covering `-i`, `-o`, `-m`,
`-d`, `-b`, `-g`, `-t`, `--no-csf`, and `--model`. Read it next to
this vignette to see the moving parts in their natural habitat.

## Avoid these traps

A few things the linter will catch but are easier to get right the
first time:

- **Don't put `%` placeholders on `flag` types.** A flag is presence /
  absence — its argstr is the literal switch (`-g`, not `-g %s`).
- **Don't redirect with shell metacharacters.** `"argstr": "> %s"` is
  illegal because niflowr never invokes a shell. Use
  `cli.stdout_to: "out_file"` instead — the runner handles redirection
  itself.
- **Don't use the same `position` for two required inputs.** Use distinct
  integers, or negative positions for trailing args.
- **Validate at the boundary.** Numeric ranges, regex patterns, and
  `exists: true` are far cheaper than a half-completed run that crashes
  inside the binary.

## Next steps

- The schema lives at `inst/schema/niflowr-spec-0.1.0.json` — every
  field above is documented there.
- Existing specs in `inst/specs/` are the best reference. `fsl.bet.json`
  shows `xor` constraints in anger; `fastsurfer.run.json` shows
  `template` output paths; `freesurfer.synthmorph_register.json`
  shows trailing positionals via negative `position`.
- See `vignette("niflowr")` for the user-facing workflow, including
  container profiles and `{targets}` integration.
