---
title: "Getting started with niflowr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with niflowr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(niflowr)
```

## Why niflowr?

Neuroimaging pipelines call many command-line tools — FSL's `bet`, ANTs'
`antsRegistration`, FreeSurfer's `recon-all` — each with its own argument
syntax, validation quirks, and output conventions. Wrapping these in R
typically means writing brittle `system()` calls with `paste()`-assembled
shell strings, no input validation, and no record of what ran.
niflowr takes a different approach:

- **Tool interfaces are defined as JSON specs**, not executable code. Each
  spec declares inputs, outputs, CLI rendering rules, and validation
  constraints.
- **Commands are built as argument vectors**, never shell strings — no
  quoting bugs, no injection hazards.
- **Execution captures provenance** — what command ran, with what arguments,
  how long it took, and what it produced.
- **Pipeline orchestration is left to {targets}** — niflowr handles the
  interface layer; {targets} and {crew} handle scheduling, caching, and
  parallelism.

## Quick example

Here's what it looks like to run FSL's brain extraction tool:

```{r quick-example, eval = FALSE}
result <- ni_fsl_bet(
  in_file  = "sub-01_T1w.nii.gz",
  out_file = "sub-01_desc-brain_T1w.nii.gz",
  frac     = 0.5,
  mask     = TRUE
)

ni_outputs(result)
#> $out_file
#> [1] "sub-01_desc-brain_T1w.nii.gz"
#> $mask_file
#> [1] "sub-01_desc-brain_T1w.nii.gz_mask.nii.gz"
```

That single call validated your inputs, built a safe argument vector,
ran `bet` via `processx`, checked that outputs exist, and wrote a
provenance sidecar — all without a single `paste()` or `system()`.

## Browsing available specs

niflowr ships with specs for common tools across four major suites:

```{r list-specs}
ni_spec_list()
```

Each spec is a JSON file under `inst/specs/`. You can inspect any spec
to see what inputs and outputs it defines:

```{r read-spec}
spec <- ni_spec_read("fsl.bet")
spec
```

## The core workflow

niflowr has three core objects that flow into each other:

1. **`ni_spec`** — the tool interface definition (from JSON)
2. **`ni_call`** — a spec bound to concrete parameter values
3. **`ni_result`** — the execution output with runtime info and provenance

### Step 1: Build a call

`ni_call()` binds parameter values to a spec. It validates inputs
immediately — required parameters, file existence, numeric ranges,
mutual exclusion constraints:
```{r build-call, error = TRUE}
# This fails: frac must be between 0 and 1
ni_call("fsl.bet",
  in_file  = "/tmp/t1.nii.gz",
  out_file = "/tmp/brain.nii.gz",
  frac     = 1.5,
  .validate = FALSE  # skip file-exists check for this demo
)
```

When inputs are valid, you get back a call object you can inspect:
```{r valid-call}
call <- ni_call("fsl.bet",
  in_file  = "/tmp/t1.nii.gz",
  out_file = "/tmp/brain.nii.gz",
  frac     = 0.3,
  mask     = TRUE,
  .validate = FALSE
)
call
```

### Step 2: Inspect the command

Before running anything, you can see exactly what command would be
executed. niflowr never hides what it's doing:

```{r inspect-cmd}
ni_cmd(call)
```

The `args` vector is what gets passed to `processx::run()` — each
element is a separate argument, never concatenated into a shell string.

### Step 3: Execute

`ni_run()` runs the command and returns an `ni_result`:

```{r execute, eval = FALSE}
result <- ni_run(call)
result
#> -- ni_result: fsl.bet [success] --
#> Exit status: 0
#> Duration: 12.3s
#>
#> -- Outputs --
#>   out_file: /tmp/brain.nii.gz
#>   mask_file: /tmp/brain.nii.gz_mask.nii.gz
```

### Dry-run mode

If you just want to preview the command without executing it, use
`dry_run = TRUE` or the shorthand `ni_dry_run()`:

```{r dry-run}
ni_dry_run("fsl.bet",
  in_file  = "/tmp/t1.nii.gz",
  out_file = "/tmp/brain.nii.gz",
  frac     = 0.5,
  .engine  = "native"
)
```

## Using the convenience wrappers

Every bundled spec has a corresponding wrapper function with typed
arguments and documentation. These are the recommended way to call tools:

| Function | Tool |
|---|---|
| `ni_fsl_bet()` | FSL BET (brain extraction) |
| `ni_fsl_flirt()` | FSL FLIRT (linear registration) |
| `ni_fsl_applywarp()` | FSL applywarp (apply warp fields) |
| `ni_afni_3dcalc()` | AFNI 3dcalc (voxelwise math) |
| `ni_afni_3dresample()` | AFNI 3dresample (reslicing) |
| `ni_ants_registration()` | ANTs registration (SyN) |
| `ni_ants_apply_transforms()` | ANTs antsApplyTransforms |
| `ni_freesurfer_recon_all()` | FreeSurfer recon-all |
| `ni_freesurfer_mri_convert()` | FreeSurfer mri_convert |

Each wrapper calls `ni_call()` + `ni_run()` internally, so you get
the same validation, safe execution, and provenance:

```{r wrapper-example, eval = FALSE}
# Linear registration with FLIRT
result <- ni_fsl_flirt(
  in_file         = "func_mean.nii.gz",
  reference       = "MNI152_T1_2mm.nii.gz",
  out_file        = "func_mean_mni.nii.gz",
  out_matrix_file = "func2mni.mat",
  dof             = 12,
  cost            = "corratio"
)
```

## Provenance

Every successful run writes a JSON sidecar next to the primary output.
This records the exact command, arguments, timing, and (optionally)
input file hashes:

```{r provenance, eval = FALSE}
# After running bet...
prov <- ni_provenance(result)
prov$command
#> [1] "bet"
prov$args
#> [1] "/data/sub-01_T1w.nii.gz" "/data/sub-01_brain.nii.gz" "-f" "0.5" "-m"
prov$duration_secs
#> [1] 12.3

# Or read a sidecar from disk
ni_provenance_read("sub-01_brain_provenance.json")
```

## Container support

niflowr can run tools inside Docker or Apptainer containers.
Container execution is profile-driven via `niflowr.yml`, with stable
container paths (`/in`, `/out`, `/work`):

```{r container, eval = FALSE}
# Configure once per project (or rely on automatic niflowr.yml discovery):
ni_config(config_file = "niflowr.yml")

result <- ni_fsl_bet(
  in_file    = "/data/t1.nii.gz",
  out_file   = "/data/brain.nii.gz",
  .engine    = "auto",   # native first, then container fallback by runtime.prefer
  .profile   = "fsl"     # maps to profiles.fsl in niflowr.yml
)
# Runs natively if `bet` exists; otherwise uses configured container backend.
```

Wrappers always accept host paths; path mapping into container paths is handled
inside `ni_run()`.

### FastSurfer: GPU-accelerated segmentation

niflowr includes specs for [FastSurfer](https://deep-mi.org/research/fastsurfer/),
a GPU-accelerated alternative to FreeSurfer that runs brain segmentation in
~5 minutes. FastSurfer runs primarily via container, so configure a profile
in `niflowr.yml`:

```yaml
profiles:
  fastsurfer:
    docker_image: "deepmi/fastsurfer:cpu-v2.4.2"
    apptainer_uri: "docker://deepmi/fastsurfer:cpu-v2.4.2"
```

Then call the wrapper:

```{r fastsurfer-example, eval = FALSE}
# Full pipeline: segmentation + surface reconstruction
result <- ni_fastsurfer_run(
  t1     = "sub-01_T1w.nii.gz",
  sid    = "sub-01",
  sd     = "derivatives/fastsurfer",
  device = "cuda",
  .engine = "docker"
)

# Segmentation only (~5 min)
seg <- ni_fastsurfer_segment(
  t1     = "sub-01_T1w.nii.gz",
  sid    = "sub-01",
  sd     = "derivatives/fastsurfer",
  device = "cpu"
)
ni_outputs(seg)
#> $seg_file
#> [1] "derivatives/fastsurfer/sub-01/mri/aparc.DKTatlas+aseg.deep.mgz"
#> $mask_file
#> [1] "derivatives/fastsurfer/sub-01/mri/mask.mgz"
```

### Runtime diagnostics with `ni_doctor()`

Use `ni_doctor()` to check runtime binaries, mounted roots, profile definitions,
and lockfile consistency:

```{r doctor, eval = FALSE}
report <- ni_doctor()
report

# Strict mode errors on failures:
ni_doctor(strict = TRUE)
```

### Pinning and validating lockfiles

Use `ni_pin()` to create/update a lockfile with resolved runtime references
for configured profiles, and `ni_lock_validate()` to verify current config
matches the lock:

```{r lockfile, eval = FALSE}
# Create/update lockfile (default: runtime.lockfile or niflowr.lock.yml)
lock <- ni_pin()

# Validate current config against lockfile
checks <- ni_lock_validate()
checks
```

### Lock enforcement during execution

Enable lock enforcement to ensure containerized runs use lockfile-pinned
references:

```{r lock-enforce, eval = FALSE}
ni_config(config = list(
  runtime = list(
    lockfile = "niflowr.lock.yml",
    lock_enforce = TRUE
  )
))

# When lock_enforce = TRUE, ni_run() resolves container refs from the lockfile
# for the selected profile and errors if lock constraints are violated.
result <- ni_fsl_bet(
  in_file  = "/data/t1.nii.gz",
  out_file = "/data/brain.nii.gz",
  .engine  = "docker",
  .profile = "fsl"
)
```

## Ecosystem integration

niflowr is designed to work with three companion packages:

### neuroim2 — read outputs as R objects

After running a tool, load its output directly into a neuroim2
`NeuroVol` or `NeuroVec`:

```{r neuroim2, eval = FALSE}
result <- ni_fsl_bet(in_file = "t1.nii.gz", out_file = "brain.nii.gz")
brain <- ni_read_output(result, "out_file")
# Returns a NeuroVol — you can slice, plot, do math on it
```

### neurotransform — work with spatial transforms

Registration outputs can be loaded as neurotransform morphism objects
for composition and resampling:

```{r neurotransform, eval = FALSE}
reg <- ni_fsl_flirt(
  in_file = "func.nii.gz", reference = "t1.nii.gz",
  out_file = "func_t1.nii.gz", out_matrix_file = "func2t1.mat"
)
xfm <- ni_read_transform(reg, "out_matrix_file")
# Returns an Affine3DMorphism — compose, invert, resample with it
```

### bidser — BIDS-aware input discovery

Query a BIDS dataset for inputs and generate derivatives-style output
paths:

```{r bidser, eval = FALSE}
library(bidser)
proj <- bids_project("/data/bids")

# Find T1w images for a subject
inputs <- ni_bids_inputs(proj, "fsl.bet", subid = "01", modality = "T1w")

# Generate a BIDS-derivatives output path
ni_deriv_path(inputs$path[1], desc = "brain", suffix = "T1w")
#> [1] "derivatives/niflowr/sub-01/anat/sub-01_desc-brain_T1w.nii.gz"
```

## Building pipelines with {targets}

For multi-step pipelines, niflowr integrates with {targets} for
DAG-based execution with caching and parallelism. The key idea:
niflowr handles the *interface layer*, {targets} handles the
*execution layer*.

### Using `ni_run_files()` in targets

The simplest pattern uses `ni_run_files()`, which returns output paths
suitable for `format = "file"` targets:

```{r targets-basic, eval = FALSE}
# _targets.R
library(targets)
library(niflowr)

list(
  tar_target(
    brain,
    ni_run_files("fsl.bet",
      in_file  = "data/sub-01_T1w.nii.gz",
      out_file = "derivatives/sub-01_desc-brain_T1w.nii.gz",
      frac = 0.5
    ),
    format = "file"
  )
)
```

### Using `tar_ni_step()` for less boilerplate

The `tar_ni_step()` target factory wraps this pattern:

```{r targets-factory, eval = FALSE}
list(
  tar_ni_step(brain, "fsl.bet",
    in_file  = "data/sub-01_T1w.nii.gz",
    out_file = "derivatives/sub-01_desc-brain_T1w.nii.gz",
    frac = 0.5
  )
)
```

### Dynamic branching over subjects

Process multiple subjects in parallel with {targets} dynamic branching
and {crew} workers:

```{r targets-branch, eval = FALSE}
library(targets)
library(crew)
library(niflowr)
library(bidser)

tar_option_set(
  packages = c("niflowr", "bidser"),
  controller = crew_controller_local(workers = 8)
)

list(
  tar_target(bids, bidser::bids_project("data/bids")),

  tar_target(t1w_files,
    bidser::search_files(bids, regex = "_T1w\\.nii", full_path = TRUE)
  ),

  tar_target(brain, {
    out <- ni_deriv_path(t1w_files, desc = "brain")
    ni_run_files("fsl.bet", in_file = t1w_files, out_file = out, frac = 0.5)
  }, pattern = map(t1w_files), format = "file")
)
```

{targets} skips subjects whose outputs are already up to date, and
{crew} distributes the work across 8 parallel workers.

## Writing your own specs

You can write specs for any command-line tool. Create a JSON file
following the schema at `inst/schema/niflowr-spec-0.1.0.json`:

```json
{
  "spec_version": "0.1.0",
  "id": "my.tool",
  "title": "My Custom Tool",
  "command": "mytool",
  "inputs": {
    "in_file": {
      "type": "file",
      "required": true,
      "validate": { "exists": true },
      "cli": { "argstr": "--input %s" }
    },
    "threshold": {
      "type": "double",
      "validate": { "min": 0, "max": 1 },
      "cli": { "argstr": "--thresh %g" }
    }
  },
  "outputs": {
    "out_file": {
      "type": "file",
      "path": { "template": "{in_file}_processed.nii.gz" },
      "must_exist": true
    }
  }
}
```

Load it by path:

```{r custom-spec, eval = FALSE}
spec <- ni_spec_read("/path/to/my.tool.json")
result <- ni_run(ni_call(spec, in_file = "data.nii.gz", threshold = 0.5))
```

## Next steps

- Browse the bundled specs with `ni_spec_list()` and `ni_spec_read()`
  to see how real interfaces are defined
- See `?ni_call` for the full set of validation and constraint options
- See `?ni_run` for execution options (echo, provenance, container)
- See `?tar_ni_step` for {targets} integration details
- See `?ni_read_output` and `?ni_read_transform` for ecosystem helpers
