---
title: From OpenNeuro to analysis with niflowr
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{From OpenNeuro to analysis with niflowr} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates an end-to-end neuroimaging workflow in R by combining two packages:

- **openneuroR**: fetches BIDS datasets from OpenNeuro
- **niflowr**: processes neuroimaging data with FSL, AFNI, ANTs, and other tools

Together, they provide a reproducible pipeline from data acquisition to analysis, entirely within R.

## Setup

Install both packages from GitHub:

```{r}
# install.packages("pak")
pak::pak("bbuchsbaum/openneuroR")
pak::pak("bbuchsbaum/niflowr")

library(openneuroR)
library(niflowr)
```

## Step 1: Browse OpenNeuro

Use `on_search()` to find datasets by keyword, then `on_dataset()` to view metadata:

```{r}
# Search for flanker task datasets
results <- on_search("flanker")

# View details for ds000102 (flanker task)
dataset <- on_dataset("ds000102")
dataset$description
```

## Step 2: Download with ni_from_openneuro()

The convenience function `ni_from_openneuro()` downloads a dataset and returns a BIDS object:

```{r}
# Download first 3 subjects
bids <- ni_from_openneuro("ds000102", subjects = c("01", "02", "03"))

# Inspect the BIDS structure
print(bids)
```

This downloads to `~/openneuro_data/ds000102/` by default. Specify `target_dir` to customize.

## Step 3: Process with niflowr

### Skull strip T1w images

Use `ni_bids_inputs()` to find files matching a specific modality, then process them:

```{r}
# Find T1w images for subject 01
inputs <- ni_bids_inputs(bids, "fsl.bet", subid = "01", modality = "T1w")

# Skull strip with FSL BET
result <- ni_fsl_bet(
  in_file = inputs$path[1],
  out_file = ni_deriv_path(inputs$path[1], desc = "brain"),
  frac = 0.5
)
```

### Register to MNI space

Chain operations by using output from one step as input to the next:

```{r}
# Register brain-extracted T1w to MNI152
reg_result <- ni_fsl_flirt(
  in_file = result$out_file,
  ref_file = ni_fsl_template("MNI152_T1_2mm_brain"),
  out_file = ni_deriv_path(result$out_file, space = "MNI152"),
  out_matrix_file = ni_deriv_path(result$out_file, suffix = "xfm", ext = "mat")
)
```

## Step 4: Scale with targets

Use `targets` to process multiple subjects in parallel:

```{r}
library(targets)

# _targets.R
tar_option_set(packages = c("niflowr", "openneuroR"))

list(
  tar_target(bids, ni_from_openneuro("ds000102", subjects = c("01", "02", "03"))),

  tar_target(
    subjects,
    bids$participants$participant_id
  ),

  tar_target(
    brain_extracted,
    {
      inputs <- ni_bids_inputs(bids, "fsl.bet", subid = subjects, modality = "T1w")
      ni_fsl_bet(
        in_file = inputs$path[1],
        out_file = ni_deriv_path(inputs$path[1], desc = "brain"),
        frac = 0.5
      )
    },
    pattern = map(subjects)
  ),

  tar_target(
    registered,
    {
      ni_fsl_flirt(
        in_file = brain_extracted$out_file,
        ref_file = ni_fsl_template("MNI152_T1_2mm_brain"),
        out_file = ni_deriv_path(brain_extracted$out_file, space = "MNI152")
      )
    },
    pattern = map(brain_extracted)
  )
)
```

Run with `targets::tar_make()`.

## Working with fMRIPrep derivatives

If the dataset has preprocessed derivatives on OpenNeuro, download and use them directly:

```{r}
# Download fMRIPrep derivatives
on_download_derivatives(
  dataset = "ds000102",
  derivative = "fmriprep",
  dest_dir = "~/openneuro_data/ds000102/derivatives",
  subjects = c("01", "02")
)

# Load confounds
confounds <- ni_fmriprep_confounds(
  "~/openneuro_data/ds000102/derivatives/fmriprep/sub-01/func/sub-01_task-flanker_desc-confounds_timeseries.tsv"
)

# Load preprocessed BOLD
preproc <- ni_fmriprep_preproc(
  "~/openneuro_data/ds000102/derivatives/fmriprep/sub-01/func/sub-01_task-flanker_space-MNI152NLin2009cAsym_desc-preproc_bold.nii.gz"
)
```

## Tips

**Caching**: OpenNeuro downloads are cached. Re-running `ni_from_openneuro()` with the same arguments will reuse existing files.

**Offline work**: Once downloaded, all processing happens locally. Disconnect and continue working.

**Reproducibility**: Use `renv::snapshot()` to lock package versions and ensure reproducible results across environments.

**Container profiles**: Use `ni_set_runtime_profile()` to switch between Docker, Apptainer, or native execution. See `vignette("container-profiles")` for details.

## Next steps

- Explore wrapper functions with `ni_list_wrappers()`
- Build complex pipelines with `ni_pipe()`
- Read `vignette("niflowr-basics")` for more processing examples
